Attribution of undesirable character traits, rather than trait-based dehumanization, predicts punishment decisions

### 2024-07-03 manuscript.Rmd
Update manuscript.Rmd

#If you want to look at a scecific number of observations with the highest Cook’s distance to assess them further:
model1A.diag.metrics %>%
  top_n(9, wt = .cooksd)

@@

#If you want to look at a spcecific number of observations with the highest Cook’s distance to assess them further:
model1B.diag.metrics %>%
  top_n(9, wt = .cooksd)

@@

plot1 <- ggplot(d.plot_1,aes(y=Harm_mean,x=Rating,color=Desirability))+
  geom_point(size=.5, alpha = .6)+
  geom_smooth(method="lm", se=TRUE, fullrange=FALSE, level=0.95, aes(fill=Desirability), size = .5, alpha = .4) +
  scale_y_continuous(name = 'Harshness of punishment endorsed (0-100)', breaks = seq(0, 100, 20), limits = c(-5, 100)) +
  scale_x_continuous(name = 'Trait attribution (0-100)', breaks = seq(0, 100, 20), limits = c(0, 100)) + 
  theme_light() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
                        legend.position = "none", legend.text = element_text(size=14), legend.title = element_text(size = 14)) +
  theme(axis.text=element_text(size = 14, colour = "black"), axis.title=element_text(size = 14)) + 
  facet_grid(switch = "y", ~ Study ~ Humanness_dimension ) +
  theme(strip.text = element_text(size = 14), strip.background = element_rect(fill = 'grey55')) +
  scale_fill_manual(values = c('#1A85FF', '#E64B35FF')) +
  scale_color_manual(values = c('#1A85FF', '#E64B35FF')) 
plot1
ggsave('figures/plot1.pdf', plot1, width = 7.5, height = 7.5, dpi = 300)

@@

Address for correspondence: Robert A. Brennan, [rb1733\@york.ac.uk](rb1733@york.ac.uk)

@@

The construct of dehumanization is often invoked to explain intergroup harm. In key evidence for this claim, @Bastian2013 reported that the extent to which participants dehumanized criminals by denying them uniquely human character traits such as refinement, rationality and morality predicted the severity of the punishment endorsed for them. We revisit this influential finding across six highly powered and preregistered studies. In Studies 1A and 1B, we conceptually replicate Bastian and colleagues' effects, demonstrating that our method is sensitive to detecting relationships between trait-based dehumanization and punishment should they occur. In subsequent studies, we investigate whether the apparent relationship between trait-based dehumanization and punishment is driven by the desirability of the traits incorporated into the stimulus set, their perceived humanness, or both. Whereas Bastian et al. asked participants to rate the extent to which criminals possessed uniquely human traits that were socially desirable, we asked participants to rate the extent to which criminals possessed uniquely human traits that were both socially desirable (e.g., cultured, civilised) and socially undesirable (e.g., arrogant, bitter). Correlational (studies 2A and 2B) and experimental (studies 3A and 3B) evidence converge on the conclusion that apparent evidence for the relationship between trait-based dehumanization and punishment is better explained by the extent to which participants attribute socially desirable attributes to criminals rather than the extent to which they attribute uniquely human attributes to criminals. Taken together, these studies cast doubt on the hypothesised causal relationship between trait-based dehumanization and harm, at least in this context.

@@

Understanding the motivations that lead to intergroup harm has been a driving force behind social psychology since its conception [@Farr1996; @Gaines1995]. Many psychological processes have been shown to exacerbate outgroup derogation, including prejudicial attitudes and stereotyping [@Dovidio2010; @Major2009]. Over the last 25 years, an increasing body of research has investigated the extent to which a psychological process of dehumanization increases the risk of intergroup harm [@Bloom2022; @Giner-Sorolla2021; @Goldenberg2021; @Lang2010; @Lang2020; @Manne2018; @Over2020a; @Over2020b; @Rai2017; @Ruiter2022; @Smith2011; @Smith2020; @Smith2021; @Vaes2020]. According to the dehumanization hypothesis, when members of an outgroup are perceived as less human than ingroup members, they are at greater risk of harm [@Haslam2019; @Haslam2014; @Haslam2015; @Smith2011; @Smith2016]. Subtle forms of dehumanization are thought to be pervasive in contemporary society. For example, the dehumanization of national groups [@Bain2011; @Leyens2003], religious groups [@Banton2020; @Leidner2013], individuals on a low income [@Loughnan2013], and refugees [@Bruneau2018; @Esses2008] has been reported in the literature.

@@

@Bastian2013 conducted one of the most prominent empirical studies testing the hypothesised association between the denial of human character traits and harm. The researchers measured how trait-based dehumanization influenced participants' punishment of criminals. Participants were asked to rate their agreement with four items assessing animalistic dehumanization of criminals: "*I felt like the person in the story was refined and cultured*" [reversed], "*I felt like the person in the story was rational and logical, like they were intelligent*" [reversed] "*I felt like the person in the story lacked self-restraint, like an animal*", and "*I felt like the person in the story was unsophisticated*". Participants were also asked to rate their agreement with four items assessing mechanistic dehumanization of criminals: "*I felt like the person in the story was open minded, like they could think clearly about things*" [reversed], "*I felt like the person in the story was emotional, like they were responsive and warm*" [reversed], "*I felt like the person in the story was superficial like they had no depth*", "*I felt like the person in the story was mechanical and cold, like a robot*". Bastian and colleagues reported that both forms of dehumanization predicted endorsement of harsh punishment for the criminals portrayed in their stimuli, concluding that their participants viewed criminals as '*subhuman and beastly*' (pp.9).

Recently, however, the explanatory value of the dual model has been called into question [@Enock2021a; @Over2020a; -@Over2020b]. According to these critiques, evidence for trait-based dehumanization is often confounded with social desirability. In Bastian and colleagues' work, evidence that criminals were animalistically dehumanized was drawn from the observation that participants judged them to be unsophisticated, lacking self-restraint, unrefined, uncultured, irrational, and unintelligent. Evidence that criminals were mechanistically dehumanized came from the observation that participants viewed them as superficial, cold, and lacking in warmth and responsiveness. These results may reflect dehumanization because the traits criminals were found to lack are those perceived as uniquely or essentially human [@Haslam2006; @Haslam2004]. However, as the traits deemed uniquely humans were all socially desirable, it is impossible to separate evidence for trait-based dehumanization from evidence for negative evaluation more generally. An alternative explanation for the findings from Bastian and colleagues is that participants endorse harsh punishment against criminals to the extent they perceive criminals to possess undesirable or antisocial characteristics.

One way to de-confound evidence for trait-based dehumanization from evidence of negative evaluation is to ask participants to rate the target group on traits that are uniquely human but socially undesirable [@Over2020a; -@Over2020b]. Previous research has established that undesirable character traits such as jealousy, spite and bitterness are considered unique to humans and socially undesirable [@Enock2021a]. Enock and colleagues [-@Enock2021a] found that participants tend to attribute socially desirable human traits more strongly to the ingroup, and socially undesirable traits more strongly to the outgroup. This result held across three intergroup contexts [see also @Enock2021b; @Enock2022]. Enock and colleagues conclude that apparent evidence for trait-based dehumanization may be better explained by intergroup preference. However, it is not yet clear how the attribution of uniquely human character traits relates to harm.

@@

All studies received ethical approval from the Psychology Departmental Ethics Committee University of York. All data collection occurred online, and the studies were created and administered using Qualtrics (https://www.qualtrics.com). Participants were recruited through the online platform Prolific (https://www.prolific.co), with an independent sample recruited for each study. Informed consent was obtained at the start of each session according to approved ethical guidelines. Inclusion criteria for each study included adult participants fluent in English who had never been to prison for committing a crime and had a Prolific approval rating of at least 90% (95% for Studies 3A and 3B). Increases in Prolific's recommended rate of compensation for participation during data collection meant the reward ranged from approximately £7 per hour in Studies 1A and 1B to approximately £8 in the other four studies. Assumption testing and analyses were conducted using *SPSS* and *RStudio*. All studies were pre-registered on AsPredicted.com before commencing data collection. Links to pre-registration documents, data files [@Brennan2023], a fully computationally reproducible version of the manuscript, and supplementary materials including the stimuli used for each study can be found at: https://osf.io/d4cvp/?view_only=ccf139206e59471b979f53708595ef05.

@@

plot2 <- ggplot(d.plot_2,aes(y=Harm_mean,x=Rating,color=Desirability_condition))+
  geom_point(size=.5, alpha = .6)+
  geom_smooth(method="lm", se=TRUE, fullrange=FALSE, level=0.95, aes(fill=Desirability_condition), size = .5, alpha = .4) +
  scale_y_continuous(name = 'Harshness of punishment endorsed (0-100)', breaks = seq(0, 100, 20), limits = c(-5, 100)) +
  scale_x_continuous(name = 'Trait attribution (0-100)',breaks = seq(0, 100, 20), limits = c(0, 100)) + 
  theme_light() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
                        legend.position = "top", legend.text = element_text(size=14), legend.title = element_text(size = 14)) +
  theme(axis.text=element_text(size = 14, colour = "black"), axis.title=element_text(size = 14)) + 
  facet_grid(switch = "y", ~ Study ~ Humanness_dimension ) +
  theme(strip.text = element_text(size = 14), strip.background = element_rect(fill = 'grey55')) +
  scale_fill_manual(name = 'Trait desirability', values = c('#1A85FF', '#D41159')) +
  scale_color_manual(name  = 'Trait desirability', values = c('#1A85FF', '#D41159')) 
plot2
ggsave('figures/plot2.pdf', plot2, width = 7, height = 7, dpi = 300)

@@

Study 2A investigated whether apparent evidence for a relationship between trait-based dehumanization and endorsement of harsh punishment for violent criminals remains when controlling for the desirability of the traits. We tested this by introducing character traits perceived as uniquely human yet socially undesirable into the stimulus set [@Enock2021a; @Over2020a]. The dual model predicts that to the extent criminals are denied uniquely human character traits, they will be subjected to harsher punishment. We predict that trait desirability will moderate the relationship between human trait attribution and punishment. More specifically, we predict that the extent to which violent criminals are denied socially desirable character traits, and attributed socially undesirable character traits, will predict harsh punishment.

@@

plot_DMHARM3_Ani <- ggplot(d.plot_main, aes(x = Humanness, y = M, group = Desirability, fill = Desirability, color = Desirability)) + 
  geom_bar(stat  = 'summary', fun = 'mean', alpha = 1, position = position_dodge(.9), width = .8) +
  geom_point(position = position_jitterdodge(dodge.width = .9, jitter.width = 0.4), alpha = 0.5, size = 0.3) + 
  stat_summary(fun.data = mean_se, geom = 'errorbar', color = 'black', width = .3, position = position_dodge(.9)) + 
  scale_y_continuous(name = 'Mean support for parole (0-100)', breaks = seq(0, 100, 20), limits = c(0, 100)) + 
  theme_light() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
                        legend.position = "top", legend.text = element_text(size=14), legend.title = element_text(size = 14)) +
  theme(axis.text=element_text(size = 14, colour = "black"), axis.title=element_text(size = 14)) + 
  facet_wrap(~Study) + 
  theme(strip.text = element_text(size = 14, colour = 'white'), strip.background = element_rect(fill = 'grey55')) +
  scale_x_discrete(name = 'Trait humanness') +
  scale_fill_manual(name = 'Trait desirability', values = c('azure1', 'seashell'), labels = c("Desirable", "Undesirable")) +
  scale_color_manual(name  = 'Trait desirability', values = c('#1A85FF', '#D41159'), labels = c("Desirable", "Undesirable")) 
plot_DMHARM3_Ani
ggsave('figures/plot_DMHARM3_Ani.png', plot_DMHARM3_Ani, width = 12, height = 8, dpi = 300)

@@

plot_DMHARM3_Mech <- ggplot(d.plot_main, aes(x = Humanness, y = M, group = Desirability, fill = Desirability, color = Desirability)) + 
  geom_bar(stat  = 'summary', fun = 'mean', alpha = 1, position = position_dodge(.9), width = .8) +
  geom_point(position = position_jitterdodge(dodge.width = .9, jitter.width = 0.4), alpha = 0.5, size = 0.3) + 
  stat_summary(fun.data = mean_se, geom = 'errorbar', color = 'black', width = .3, position = position_dodge(.9)) + 
  scale_y_continuous(name = 'Mean support for parole (0-100)', breaks = seq(0, 100, 20), limits = c(0, 100)) + 
  theme_light() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
                        legend.position = "top", legend.text = element_text(size=14), legend.title = element_text(size = 14)) +
  theme(axis.text=element_text(size = 14, colour = "black"), axis.title=element_text(size = 14)) + 
  facet_wrap(~Study) + 
  theme(strip.text = element_text(size = 14, colour = 'white'), strip.background = element_rect(fill = 'grey55')) +
  scale_x_discrete(name = 'Trait humanness') +
  scale_fill_manual(name = 'Trait desirability', values = c('azure1', 'seashell'), labels = c("Desirable", "Undesirable")) +
  scale_color_manual(name  = 'Trait desirability', values = c('#1A85FF', '#D41159'), labels = c("Desirable", "Undesirable")) 
plot_DMHARM3_Mech
ggsave('figures/plot_DMHARM3_Mech.png', plot_DMHARM3_Mech, width = 12, height = 8, dpi = 300)

@@

In Studies 2A and 2B, we investigated the extent to which the previously reported relationship between trait-based dehumanization and harm can be explained by the social desirability of the traits incorporated into the stimulus set. The dual model [@Haslam2006] has previously been critiqued for failing to clearly distinguish evidence for trait-based dehumanization from evidence of negative evaluation [@Bloom2022; @Enock2021a; @Over2020a; -@Over2020b]. Bastian and colleagues [-@Bastian2013] operationalised animalistic dehumanization as a reduction in the extent to which participants viewed criminals as possessing traits such as sophistication and refinement. They operationalised mechanistic dehumanization as a reduction in the extent to which participants viewed criminals as possessing traits like warmth and depth. As each of these human traits is socially desirable, it is impossible to determine whether harm is predicted by dehumanization or negative evaluation. In order to tease apart the influence of dehumanization and negative evaluation in harm, we incorporated undesirable human traits into our stimulus set, for example, bitter and spiteful. If trait-based dehumanization explains harm, the previously reported relationship between dehumanization and punishment should remain even when undesirable human traits are incorporated into the stimulus set. If the previously reported relationship is better explained by negative evaluation, then trait desirability should moderate the relationship with punishment. In support of the latter claim, Studies 2A and 2B showed that the more desirable human traits participants attributed to criminals, the less harshly participants thought they should be punished. The more undesirable human traits participants attributed to criminals, the more harshly participants thought they should be punished.

@@

Partially in response to these critiques, @Kteily2022 presented a new social psychological model of dehumanization in which to dehumanize an individual or group is to perceive them as less than the ideal human. Under this characterisation of dehumanization, to view a group as possessing negative attributes is to dehumanize them. However, to define dehumanization in such a broad way as any negative evaluation renders almost all social judgments dehumanizing [@Bloom2022]. It seems unlikely that we dehumanize our closest and most loved kin simply by perceiving their imperfections. It is crucial that future conceptual research on dehumanization more clearly delineates dehumanization from negative evaluation [@Bloom2022; @Over2020a; -@Over2020b].

@@

We are not trying to argue that trait-based dehumanization never occurs. Rather, our argument is considerably more modest. Taken in conjunction with other recent results, it is apparent that evidence for trait-based dehumanization has often been confounded with evidence for negative evaluation [@Bloom2022; @Enock2021a; @Enock2021b; @Over2020a; -@Over2020b]. The results of the current study add to this growing body of critiques by showing that the findings of @Bastian2013, often cited as key evidence for the claim that trait-based dehumanization leads to harm, are considerably less convincing than they first appear. It is imperative that future research tests whether there is evidence for trait-based dehumanization when trait desirability is controlled for, given the grave importance of understanding predictors of intergroup harm in the real world.

>>>>

#If you want to look at a specific number of observations with the highest Cook’s distance to assess them further:
model1A.diag.metrics %>%
  top_n(9, wt = .cooksd)

@@

#If you want to look at a specific number of observations with the highest Cook’s distance to assess them further:
model1B.diag.metrics %>%
  top_n(9, wt = .cooksd)

@@

plot1 <- ggplot(d.plot_1,aes(y=Harm_mean,x=Rating,color=Desirability))+
  geom_point(size=.5, alpha = .6)+
  geom_smooth(method="lm", se=TRUE, fullrange=FALSE, level=0.95, aes(fill=Desirability), size = .5, alpha = .4) +
  scale_y_continuous(name = 'Harshness of punishment endorsed (0-100)', breaks = seq(0, 100, 20), limits = c(-5, 100)) +
  scale_x_continuous(name = 'Trait attribution (0-100)', breaks = seq(0, 100, 20), limits = c(0, 100)) + 
  theme_light() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
                        legend.position = "none", legend.text = element_text(size=14), legend.title = element_text(size = 14)) +
  theme(axis.text=element_text(size = 14, colour = "black"), axis.title=element_text(size = 14)) + 
  facet_grid(switch = "y", ~ Study ~ Humanness_dimension ) +
  theme(strip.text = element_text(size = 14), strip.background = element_rect(fill = 'grey40')) +
  scale_fill_manual(values = c('steelblue2', 'tomato')) +
  scale_color_manual(values = c('mediumblue', 'firebrick')) 
plot1
ggsave('figures/plot1.pdf', plot1, width = 7.5, height = 7.5, dpi = 300)

@@

Address for correspondence: Robert A. Brennan, [rb1733\@york.ac.uk](rb1733@york.ac.uk); [brennan.rob.a@gmail.com](brennan.rob.a@gmail.com)

@@

Previous work has reported that the extent to which participants dehumanized criminals by denying them uniquely human character traits such as refinement, rationality and morality predicted the severity of the punishment endorsed for them. We revisit this influential finding across six highly powered and pre-registered studies. First, we conceptually replicate the effect reported in previous work, demonstrating that our method is sensitive to detecting relationships between trait-based dehumanization and punishment should they occur. We then investigate whether the apparent relationship between trait-based dehumanization and punishment is driven by the desirability of the traits incorporated into the stimulus set, their perceived humanness, or both. To do this, we asked participants to rate the extent to which criminals possessed uniquely human traits that were either socially desirable (e.g. cultured and civilized) or socially undesirable (e.g. arrogant and bitter). Correlational and experimental evidence converge on the conclusion that apparent evidence for the relationship between trait-based dehumanization and punishment is better explained by the extent to which participants attribute socially desirable attributes to criminals rather than the extent to which they attribute uniquely human attributes. These studies cast doubt on the hypothesized causal relationship between trait-based dehumanization and harm, at least in this context.

@@

Understanding the motivations that lead to intergroup harm has been a driving force behind social psychology since its conception [@Farr1996; @Gaines1995]. Many psychological processes have been shown to exacerbate outgroup derogation, including prejudicial attitudes and stereotyping [@Dovidio2010; @Major2009]. Over the last 25 years, an increasing body of research has investigated the extent to which a psychological process of dehumanization increases the risk of intergroup harm [@Bloom2017; @Bloom2022; @Giner-Sorolla2021; @Goldenberg2021; @Lang2010; @Lang2020; @Manne2016; @Manne2018; @Over2021a; @Over2021b; @Rai2017; @Ruiter2022; @Smith2011; @Smith2020; @Smith2021; @Vaes2021]. According to the dehumanization hypothesis, when members of an outgroup are perceived as less human than ingroup members, they are at greater risk of harm [@Haslam2019; @Haslam2014; @Haslam2015; @Smith2011; @Smith2016]. Subtle forms of dehumanization are thought to be pervasive in contemporary society. For example, the dehumanization of national groups [@Bain2011; @Leyens2003], religious groups [@Banton2020; @Leidner2013], individuals on a low income [@Loughnan2013], and refugees [@Bruneau2018; @Esses2008] has been reported in the literature.

@@

@Bastian2013 conducted influential empirical studies testing the hypothesised association between the denial of human character traits and the endorsement of harsh punishment. The researchers measured how trait-based dehumanization influenced participants' punishment of criminals. Participants were asked to rate their agreement with four items assessing animalistic dehumanization of criminals: "*I felt like the person in the story was refined and cultured*" [reversed], "*I felt like the person in the story was rational and logical, like they were intelligent*" [reversed] "*I felt like the person in the story lacked self-restraint, like an animal*", and "*I felt like the person in the story was unsophisticated*". Participants were also asked to rate their agreement with four items assessing mechanistic dehumanization of criminals: "*I felt like the person in the story was open minded, like they could think clearly about things*" [reversed], "*I felt like the person in the story was emotional, like they were responsive and warm*" [reversed], "*I felt like the person in the story was superficial like they had no depth*", "*I felt like the person in the story was mechanical and cold, like a robot*". Bastian and colleagues reported that both forms of dehumanization predicted endorsement of harsh punishment for the criminals portrayed in their stimuli, concluding that their participants viewed criminals as '*subhuman and beastly*' (pp.9).

Recently, however, the explanatory value of the dual model has been called into question [@Enock2021a; @Over2021a; -@Over2021b]. According to these critiques, evidence for trait-based dehumanization is often confounded with social desirability. In Bastian and colleagues' work, evidence that criminals were animalistically dehumanized was drawn from the observation that participants judged them to be unsophisticated, lacking self-restraint, unrefined, uncultured, irrational, and unintelligent. Evidence that criminals were mechanistically dehumanized came from the observation that participants viewed them as superficial, cold, and lacking in warmth and responsiveness. These results may reflect dehumanization because the traits criminals were found to lack are those perceived as uniquely or essentially human [@Haslam2006; @Haslam2004]. However, as the traits deemed uniquely humans were all socially desirable, it is impossible to separate evidence for trait-based dehumanization from evidence for negative evaluation more generally. An alternative explanation for the findings from Bastian and colleagues is that participants endorse harsh punishment against criminals to the extent they perceive criminals to possess undesirable or antisocial characteristics.

One way to de-confound evidence for trait-based dehumanization from evidence of negative evaluation is to ask participants to rate the target group on traits that are uniquely human but socially undesirable [@Over2021a; -@Over2021b]. Previous research has established that undesirable character traits such as jealousy, spite and bitterness are considered unique to humans and socially undesirable [@Enock2021a]. Enock and colleagues [-@Enock2021a] found that participants tend to attribute socially desirable human traits more strongly to the ingroup, and socially undesirable traits more strongly to the outgroup. This result held across three intergroup contexts [see also @Enock2021b; @Enock2022]. Enock and colleagues conclude that apparent evidence for trait-based dehumanization may be better explained by intergroup preference. However, it is not yet clear how the attribution of uniquely human character traits relates to harm.

@@

All studies received ethical approval from the Psychology Departmental Ethics Committee at the University of York (approval number 926). All data collection occurred online, and the studies were created and administered using Qualtrics (https://www.qualtrics.com). Participants were recruited through the online platform Prolific (https://www.prolific.co), with an independent sample recruited for each study. Informed consent was obtained at the start of each session according to approved ethical guidelines. Inclusion criteria for each study included adult participants fluent in English who had never been to prison for committing a crime and had a Prolific approval rating of at least 90% (95% for Studies 3A and 3B). Increases in Prolific's recommended rate of compensation for participation during data collection meant the reward ranged from approximately £7 per hour in Studies 1A and 1B to approximately £8 in the other four studies. Assumption testing and analyses were conducted using *SPSS* and *RStudio*. All studies were pre-registered on AsPredicted.com before commencing data collection. Links to pre-registration documents, data files [@Brennan2023], a fully computationally reproducible version of the manuscript, and electronic supplementary materials including the stimuli used for each study can be found at: http://doi.org/10.17605/OSF.IO/D4CVP.

@@

plot2 <- ggplot(d.plot_2,aes(y=Harm_mean,x=Rating,color=Desirability_condition))+
  geom_point(size=.5, alpha = .6)+
  geom_smooth(method="lm", se=TRUE, fullrange=FALSE, level=0.95, aes(fill=Desirability_condition), size = .5, alpha = .4) +
  scale_y_continuous(name = 'Harshness of punishment endorsed (0-100)', breaks = seq(0, 100, 20), limits = c(-5, 100)) +
  scale_x_continuous(name = 'Trait attribution (0-100)',breaks = seq(0, 100, 20), limits = c(0, 100)) + 
  theme_light() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
                        legend.position = "top", legend.text = element_text(size=14), legend.title = element_text(size = 14)) +
  theme(axis.text=element_text(size = 14, colour = "black"), axis.title=element_text(size = 14)) + 
  facet_grid(switch = "y", ~ Study ~ Humanness_dimension ) +
  theme(strip.text = element_text(size = 14), strip.background = element_rect(fill = 'grey40')) +
  scale_fill_manual(name = 'Trait desirability', values = c('steelblue', 'tomato')) +
  scale_color_manual(name  = 'Trait desirability', values = c('mediumblue', 'firebrick')) 
plot2
ggsave('figures/plot2.pdf', plot2, width = 7, height = 7, dpi = 300)

@@

Study 2A investigated whether apparent evidence for a relationship between trait-based dehumanization and endorsement of harsh punishment for violent criminals remains when controlling for the desirability of the traits. We tested this by introducing character traits perceived as uniquely human yet socially undesirable into the stimulus set [@Enock2021a; @Over2021a]. The dual model predicts that to the extent criminals are denied uniquely human character traits, they will be subjected to harsher punishment. We predict that trait desirability will moderate the relationship between human trait attribution and punishment. More specifically, we predict that the extent to which violent criminals are denied socially desirable character traits, and attributed socially undesirable character traits, will predict harsh punishment.

@@

plot_DMHARM3_Ani <- ggplot(d.plot_main, aes(x = Humanness, y = M, group = Desirability, fill = Desirability, color = Desirability)) + 
  geom_bar(stat  = 'summary', fun = 'mean', alpha = 1, position = position_dodge(.9), width = .8) +
  geom_point(position = position_jitterdodge(dodge.width = .9, jitter.width = 0.4), alpha = 0.5, size = 0.3) + 
  stat_summary(fun.data = mean_se, geom = 'errorbar', color = 'black', width = .3, position = position_dodge(.9)) + 
  scale_y_continuous(name = 'Mean support for parole (0-100)', breaks = seq(0, 100, 20), limits = c(0, 100)) + 
  theme_light() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
                        legend.position = "top", legend.text = element_text(size=14), legend.title = element_text(size = 14)) +
  theme(axis.text=element_text(size = 14, colour = "black"), axis.title=element_text(size = 14)) + 
  facet_wrap(~Study) + 
  theme(strip.text = element_text(size = 14, colour = 'white'), strip.background = element_rect(fill = 'grey40')) +
  scale_x_discrete(name = 'Trait humanness') +
  scale_fill_manual(name = 'Trait desirability', values = c('aliceblue', 'mistyrose'), labels = c("Desirable", "Undesirable")) +
  scale_color_manual(name  = 'Trait desirability', values = c('mediumblue', 'firebrick'), labels = c("Desirable", "Undesirable")) 
plot_DMHARM3_Ani
ggsave('figures/plot_DMHARM3_Ani.png', plot_DMHARM3_Ani, width = 12, height = 8, dpi = 300)

@@

plot_DMHARM3_Mech <- ggplot(d.plot_main, aes(x = Humanness, y = M, group = Desirability, fill = Desirability, color = Desirability)) + 
  geom_bar(stat  = 'summary', fun = 'mean', alpha = 1, position = position_dodge(.9), width = .8) +
  geom_point(position = position_jitterdodge(dodge.width = .9, jitter.width = 0.4), alpha = 0.5, size = 0.3) + 
  stat_summary(fun.data = mean_se, geom = 'errorbar', color = 'black', width = .3, position = position_dodge(.9)) + 
  scale_y_continuous(name = 'Mean support for parole (0-100)', breaks = seq(0, 100, 20), limits = c(0, 100)) + 
  theme_light() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
                        legend.position = "top", legend.text = element_text(size=14), legend.title = element_text(size = 14)) +
  theme(axis.text=element_text(size = 14, colour = "black"), axis.title=element_text(size = 14)) + 
  facet_wrap(~Study) + 
  theme(strip.text = element_text(size = 14, colour = 'white'), strip.background = element_rect(fill = 'grey40')) +
  scale_x_discrete(name = 'Trait humanness') +
  scale_fill_manual(name = 'Trait desirability', values = c('aliceblue', 'mistyrose'), labels = c("Desirable", "Undesirable")) +
  scale_color_manual(name  = 'Trait desirability', values = c('mediumblue', 'firebrick'), labels = c("Desirable", "Undesirable")) 
plot_DMHARM3_Mech
ggsave('figures/plot_DMHARM3_Mech.png', plot_DMHARM3_Mech, width = 12, height = 8, dpi = 300)

@@

In Studies 2A and 2B, we investigated the extent to which the previously reported relationship between trait-based dehumanization and harm can be explained by the social desirability of the traits incorporated into the stimulus set. The dual model [@Haslam2006] has previously been critiqued for failing to clearly distinguish evidence for trait-based dehumanization from evidence of negative evaluation [@Bloom2022; @Enock2021a; @Over2021a; -@Over2021b]. Bastian and colleagues [-@Bastian2013] operationalised animalistic dehumanization as a reduction in the extent to which participants viewed criminals as possessing traits such as sophistication and refinement. They operationalised mechanistic dehumanization as a reduction in the extent to which participants viewed criminals as possessing traits like warmth and depth. As each of these human traits is socially desirable, it is impossible to determine whether harm is predicted by dehumanization or negative evaluation. In order to tease apart the influence of dehumanization and negative evaluation in harm, we incorporated undesirable human traits into our stimulus set, for example, bitter and spiteful. If trait-based dehumanization explains harm, the previously reported relationship between dehumanization and punishment should remain even when undesirable human traits are incorporated into the stimulus set. If the previously reported relationship is better explained by negative evaluation, then trait desirability should moderate the relationship with punishment. In support of the latter claim, Studies 2A and 2B showed that the more desirable human traits participants attributed to criminals, the less harshly participants thought they should be punished. The more undesirable human traits participants attributed to criminals, the more harshly participants thought they should be punished.

@@

Partially in response to these critiques, @Kteily2022 presented a new social psychological model of dehumanization in which to dehumanize an individual or group is to perceive them as less than the ideal human. Under this characterisation of dehumanization, to view a group as possessing negative attributes is to dehumanize them. However, to define dehumanization in such a broad way as any negative evaluation renders almost all social judgments dehumanizing [@Bloom2022]. It seems unlikely that we dehumanize our closest and most loved kin simply by perceiving their imperfections. It is crucial that future conceptual research on dehumanization more clearly delineates dehumanization from negative evaluation [@Bloom2022; @Over2021a; -@Over2021b].

@@

We are not trying to argue that trait-based dehumanization never occurs. Rather, our argument is considerably more modest. Taken in conjunction with other recent results, it is apparent that evidence for trait-based dehumanization has often been confounded with evidence for negative evaluation [@Bloom2022; @Enock2021a; @Enock2021b; @Over2021a; -@Over2021b]. The results of the current study add to this growing body of critiques by showing that the findings of @Bastian2013, often cited as key evidence for the claim that trait-based dehumanization leads to harm, are considerably less convincing than they first appear. It is imperative that future research tests whether there is evidence for trait-based dehumanization when trait desirability is controlled for, given the grave importance of understanding predictors of intergroup harm in the real world.

**Ethical statement.** All studies received ethical approval from the Psychology Departmental Ethics Committee at the University of York (approval number 926). Informed consent was obtained from participants at the start of each experimental session according to approved ethical procedures. All studies were performed in accordance with the relevant guidelines and regulations.

**Data accessibility.** All studies reported in this article were pre-registered and the data is available open access. Data files, pre-registration documents, a fully computationally reproducible version of the manuscript, and supplementary materials, including the stimuli used for each study, can be found at http://doi.org/10.17605/OSF.IO/D4CVP.

**Declaration of AI use.** We have not used AI-assisted technologies in creating this article.

**Authors' contributions.** R.A.B.: conceptualization, data curation, formal analysis, investigation, methodology, project administration, visualization, writing—original draft, writing—review and editing; F.E.E.: conceptualization, formal analysis, investigation, methodology, resources, supervision, validation, visualization, writing—review and editing; H.O.: conceptualization, funding acquisition, investigation, methodology, project administration, resources, supervision, writing—review and editing. All authors gave final approval for publication and agreed to be held accountable for the work performed therein.

**Conflict of interest declaration.** We declare we have no competing interests.

**Funding.** This research was supported by the European Research Council under the European Union's Horizon 2020 Programme, grant number ERC- STG-755719 awarded to HO.

**Acknowledgements.** Production of the reproducible version of this manuscript was supported by an Enhancing Research Culture award from Research England.
